---
title: "Data Visualisation with ggplot2"
keypoints:
- "`ggplot2` is a flexible and useful tool for creating plots in R."
- "The data set and coordinate system can be defined using the `ggplot` function."
- "Additional layers, including geoms, are added using the `+` operator."
- "Boxplots are useful for visualizing the distribution of a continuous variable."
- "Barplot are useful for visualizing categorical data."
- "Faceting allows you to generate multiple plots based on a categorical variable."
objectives:
- "Produce scatter plots, boxplots, and time series plots using ggplot."
- "Set universal plot settings."
- "Describe what faceting is and apply faceting in ggplot."
- "Modify the aesthetics of an existing ggplot plot (including axis labels and color)."
- "Build complex and customized plots from data in a data frame."
questions:
- "What are the components of a ggplot?"
- "How do I create scatterplots, boxplots, and barplots?"
- "How can I change the aesthetics (ex. colour, transparency) of my plot?"
- "How can I create multiple plots at once?"
teaching: 80
exercises: 35
source: Rmd
---

```{r, include=FALSE}
source("../bin/download_data.R")
source("../bin/chunk-options.R")
knitr_fig_path("04-")
library(readr)
books2 <- read_csv("./data/books_reformatted.csv")
```

## Getting set up

### Set up your directories and data

If you have not already done so, open your R Project file (`library_carpentry.Rproj`) created in the `Before We Start` lesson.

**If you did not complete that step** then do the following. Only do this if you
didn't complete it in previous lessons.

* Under the `File` menu, click on `New project`, choose `New directory`, then
`New project`
* Enter the name `library_carpentry` for this new folder (or "directory"). This
will be your **working directory** for the rest of the day.
* Click on `Create project`
* Create a new file where we will type our scripts. Go to File > New File > R
script. Click the save icon on your toolbar and save your script as
"`script.R`".
* Copy and paste the below lines of code to create three new subdirectories and
download the original and the reformatted `books` data:

```{r create-dirs, eval=FALSE}
library(fs)   # https://fs.r-lib.org/.  fs is a cross-platform, uniform interface to file system operations via R. 
dir_create("data")
dir_create("data_output")
dir_create("fig_output")
download.file("https://ndownloader.figshare.com/files/22031487",
              "data/books.csv", mode = "wb")
download.file("https://ndownloader.figshare.com/files/22031487",
              "data/books.csv", mode = "wb")
download.file("https://ndownloader.figshare.com/files/22051506",
              "data_output/books_reformatted.csv", mode = "wb")
```

### Load the `tidyverse` and data frame into your R session

Load the `tidyverse` and the `lubridate` packages. `lubridate` is installed with
the tidyverse, but is not one of the core tidyverse packages loaded with
`library(tidyverse)`, so it needs to be explicitly called. `lubridate` makes
working with dates and times easier in R.

```{r load-data, purl=FALSE, results='hold'}
library(tidyverse)  # load the core tidyverse
library(lubridate)  # load lubridate
```

We also load the `books_reformatted` data we saved in the previous
lesson. We'll assign it to `books2`.

```{r, purl=FALSE, eval=FALSE}
books2 <- read_csv("data_output/books_reformatted.csv")  # load the data and assign it to books
```

## Plotting with **`ggplot2`**

Base R contains a number of functions for quick data visualization such as
`plot()` for scatter plots, `barplot()`, `hist()` for histograms, and
`boxplot()`. However, just as data manipulation is easier with `dplyr` than Base
R, so data visualization is easier with `ggplot2` than Base R. `ggplot2` is a
[“grammar” for data visualization](http://vita.had.co.nz/papers/layered-grammar.html) also created
by Hadley Wickham, as an implementation of Leland Wilkinson’s [Grammar of Graphics](http://www.worldcat.org/oclc/934513040).

`ggplot` is a plotting package that makes it simple to create complex plots
from data stored in a data frame. It provides a programmatic interface for
specifying what variables to plot, how they are displayed, and general visual
properties. Therefore, we only need minimal changes if the underlying data
change or if we decide to change from a bar plot to a scatterplot. This helps in
creating publication quality plots with minimal amounts of adjustments and
tweaking.

`ggplot2` functions like data in the 'long' format, i.e., a column for every
dimension, and a row for every observation. Well-structured data will save you
lots of time when making figures with `ggplot2`

ggplot graphics are built step by step by adding new elements. Adding layers in
this fashion allows for extensive flexibility and customization of plots.

To build a ggplot, we will use the following basic template that can be used for different types of plots:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

- use the `ggplot()` function and bind the plot to a specific data frame using
  the `data` argument

When you run the `ggplot()` function, it plots directly to the Plots tab in the
Navigation Pane (lower right). Alternatively, you can assign a plot to an R
object, then call `print()`to view the plot in the Navigation Pane.

Let's create a `booksPlot` and limit our visualization to only items in `subCollection` general collection, juvenile, and k-12, and filter out items with `NA` in `call_class`. We do this by using the <kbd>|</kbd> key on the keyboard to specify a boolean OR, and use the `!is.na()` function to keep only those items that are NOT `NA` in the `call_class` column.

```{r, purl=FALSE}
# create a new data frame
booksPlot <- books2 %>%
  filter(subCollection == "general collection" | 
           subCollection == "juvenile" | 
           subCollection == "k-12 materials",
         !is.na(call_class))
```



**`ggplot2`**
**`ggplot2`** functions like data in the 'long' format, i.e., a column for every
dimension, and a row for every observation. Well-structured data will save you
lots of time when making figures with **`ggplot2`**

ggplot graphics are built step by step by adding new elements. Adding layers in
this fashion allows for extensive flexibility and customization of plots.

To build a ggplot, we will use the following basic template that can be used for different types of plots:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

Use the `ggplot()` function and bind the plot to a specific data frame using the
`data` argument.


```{r, purl=FALSE}
ggplot(data = booksPlot)  # a blank canvas
```

Not very interesting. We need to add layers to it by defining a mapping
aesthetic and adding `geoms`.

## Define a mapping with `aes()` and display data with `geoms`

Define a mapping (using the aesthetic (`aes()`) function), by selecting the
variables to be plotted and specifying how to present them in the graph, e.g. as
x/y positions or characteristics such as size, shape, color, etc.

```{r, purl=FALSE}
ggplot(data = booksPlot, mapping = aes(x = call_class)) # define the x axis aesthetic
```

Here we define the x axes, but because we have not yet added any `geoms`, we still
do not see any data being visualized.

Data is visualized in the canvas with “geometric shapes” such as bars and lines;
what are called *geoms*. In your console, type `geom_` and press the tab key to
see the geoms included–there are over 30. For example:

* `geom_point()` for scatter plots, dot plots, etc.
* `geom_boxplot()` for boxplots
* `geom_bar()` for barplots
* `geom_line()` for trend lines, time series, etc.

Each geom takes a `mapping` argument within the `aes()` call. This is called the
*aesthetic mapping argument*. In other words, inside the geom function is an
`aes()` function, and inside `aes()` is a mapping argument specifying how to map
the variables inside the visualization. ggplot then looks for that variable
inside the data argument, and plots it accordingly.

For example, in the below expression, the `call_class` variable is being mapped
to the `x` axis in the geometric shape of a bar. In this example, the y axis
(count) is not specified, nor is it a variable in the original dataset, but is
the result of geom_bar() *binning* your data inside each call number class and
plotting the bin counts (the number of items falling into each bin).

To add a geom to the plot use the `+` operator.

```{r, purl=FALSE}
# add a bar geom and set call_class as the x axis
ggplot(data = booksPlot, mapping = aes(x = call_class)) +
  geom_bar()
```

We can see that there are about 1,500 books in the H class, 1,000 books in the P
class, 700 books in the E class, etc. As we shall see, the number of E and F
books are deceptively large, as easy books and fiction books both begin with E
and are thus lumped into that category, though they are not of the E and F
Library of Congress call number classification.

> ## `ggplot` tips
>
> - Anything you put in the `ggplot()` function can be seen by any geom layers
>   that you add (i.e., these are universal plot settings). This includes the x-
>   and y-axis mapping you set up in `aes()`.
> - You can also specify mappings for a given geom independently of the mapping
>   defined globally in the `ggplot()` function.
> - The `+` sign used to add new layers must be placed at the end of the line
>   containing the *previous* layer. If, instead, the `+` sign is added at the
>   beginning of the line containing the new layer, **`ggplot2`** will not add
>   the new layer and will return an error message.
{: .callout}

## Univariate geoms

*“Univariate”* refers to a single variable. A histogram is a univariate plot: it
shows the frequency counts of each value inside a single variable. Let’s say we
want to visualize a frequency distibution of checkouts in the `booksPlot` data. In
other words, how many items have 1 checkout? How many have 2 checkouts? And so
on.

```{r, purl=FALSE}
ggplot(data = booksPlot, mapping = aes(x = tot_chkout)) +
  geom_histogram()
```

As we have seen in previous lessons, the overwhelming majority of books have a
small amount of usage, so the plot is heavily skewed. As anyone who has done
collection analysis has encountered, this is a very common issue. It can be
addressed in two ways:

First, add a `binwidth` argument to `aes()`. In a histogram, each bin contains
the number of occurrences of items in the data set that are contained within
that bin. As stated in the documentation for `?geom_histogram`, "You should
always override this value, exploring multiple widths to find the best to
illustrate the stories in your data."

Second, change the scales of the `y` axes by adding another argument to ggplot:

```{r, purl=FALSE}
ggplot(data = booksPlot) +
  geom_histogram(aes(x = tot_chkout), binwidth = 10) +
  scale_y_log10()
```

Notice the scale y axis now goes from 0-10, 10-100, 100-1000, and 1000-10000.
This is called “logarithmic scale” and is based on orders of magnitude. We can
therefore see that over 5,000 books (on the y axis) have between 0-10 checkouts
(on the x axis), 1,000 books have 10-20 checkouts, and further down on the x
axis, a handful of books have 60-70 checkouts, and a handful more have around
100 checkouts.

We can check this with `table()`:

```{r, purl=FALSE}
table(booksPlot$tot_chkout)
```

ggplot has thus given us an easy way to visualize the distribution of checkouts.
If you test this on your own print and ebook usage data, you will likely find
something similar.

### Changing the geom

This same exact data can be visualized in a couple different ways by replacing the geom_histogram() function with either `geom_density()` (adding a logarithmic x scale) or `geom_freqpoly()`:

```{r, purl=FALSE}
# create a density plot
ggplot(data = booksPlot) +
  geom_density(aes(x = tot_chkout)) +
  scale_y_log10() +
  scale_x_log10()

# create a frequency polygon
ggplot(data = booksPlot) +
  geom_freqpoly(aes(x = tot_chkout), binwidth = 30) +
  scale_y_log10()
```

## Bivariate geoms

Bivariate plots visualize two variables. Let's take a look at some higher usage
items, but first eliminate the `NA` values and keep only items with more than 10
checkouts, which we will do with `filter()` from the `dplyr` package and assign
it to `booksHighUsage`

```{r}
# filter booksPlot to include only items with over 10 checkouts
booksHighUsage <- booksPlot %>%
  filter(!is.na(tot_chkout),
                tot_chkout > 10)
```

We then visualize checkouts by call number with a scatter plot. There is
still so much skew that I retain the logarithmic scale on the y axis with
`scale_y_log10()`.

```{r}
# scatter plot high usage books by call number class
ggplot(data = booksHighUsage,
       aes(x = call_class, y = tot_chkout)) +
  geom_point() +
  scale_y_log10()
```

Again, notice the scale on the y axis. We can obseve a few items of interest here: No items in the D, J, M, and Z class have more than 30 checkouts. An item in the E class has the most checkouts with over 100, but, as noted above, this includes Easy books classified with `E`, not just items with Library of Congress `E` classification (United States history) an issue we'll look at further down.

Just as with univariate plots, we can use different geoms to view various aspects of the data, which in turn reveal different patterns.

```{r}
# boxplot plot high usage books by call number class
ggplot(data = booksHighUsage,
       aes(x = call_class, y = tot_chkout)) +
  geom_boxplot() +
  scale_y_log10()
```

By adding points to a boxplot, we can have a better idea of the number of
measurements and of their distribution. Here we set the boxplot `alpha` to `0`,
which will make it see-through. We also add another layer of geom to the plot
called `geom_jitter()`, which will introduce a little bit of randomness into the
position of our points. We set the `color` of these points to `"tomato"`.

```{r boxplot-with-points, purl=FALSE}
ggplot(data = booksHighUsage, aes(x = call_class, y = tot_chkout)) +
  geom_boxplot(alpha = 0) +
  geom_jitter(alpha = 0.5, color = "tomato") +
  scale_y_log10()
```

Notice how the boxplot layer is behind the jitter layer? What do you need to
change in the code to put the boxplot in front of the points such that it's not
hidden?

> ## Plotting Exercise
>
> Boxplots are useful summaries, but hide the *shape* of the distribution. For
> example, if the distribution is bimodal, we would not see it in a
> boxplot. An alternative to the boxplot is the violin plot, where the shape
> (of the density of points) is drawn.
>
> - Replace the box plot with a violin plot; see `geom_violin()`.
>
> > ## Solution
> >
> > ```{r violin-plot}
> > ggplot(data = booksHighUsage, aes(x = call_class, y = tot_chkout)) +
> >   geom_violin(alpha = 0) +
> >   geom_jitter(alpha = 0.5, color = "tomato")
> > ```
> {: .solution}
>
> - Add color to the `geom_jitter` argument on your boxplot according to the item `subCollection`. *Hint:* If you get the error `object 'subCollection' not found` or `invalid color name 'subCollection'` then consider `color` as an aesthetic mapping.
>
> > ## Solution
> > ```{r boxplot-exercise-subcollection}
> > ggplot(data = booksHighUsage, aes(x = call_class, y = tot_chkout)) +
> > geom_violin(alpha = 0) +
> > geom_jitter(alpha = 0.5, aes(color = subCollection)) +
> > scale_y_log10()
> > ```
> {: .solution}
>
> So far, we've looked at the distribution of checkouts within call number ranges. Try
> making a new plot to explore the distribution of checkouts within another variable.
>
> - Still using the `booksHighUsage` data, create a boxplot for `tot_chkout` for each `subCollection`. Overlay the boxplot layer on a jitter layer to show actual measurements. Keep the `scale_y_log10` argument.
>
> > ## Solution
> > ```{r}
> > ggplot(data = booksHighUsage, aes(x = subCollection, y = tot_chkout)) +
> >  geom_boxplot(alpha = 0) +
> >  geom_jitter(alpha = 0.5, color = "tomato") +
> >  scale_y_log10()
> > ```
> {: .solution}
{: .challenge}


## Add a third variable

As we saw in that exercise, you can convey even more information in your
visualization by adding a third variable, in addition to the first two on the x
and y scales.

### Add a third variable with `aes()`

We can use arguments in `aes()` to map a visual aesthetic in the plot to a
variable in the dataset. Specifically, we will map `color` to the
`subCollection` variable. Because we are now mapping features of the data to a
color, instead of setting one color for all points, the color now needs to be
set inside a call to the **`aes`** function. **`ggplot2`** will provide a
different color corresponding to different values in the vector. In other words,
associate the name of the aesthetic (`color`) to the name of the variable
(`subCollection`) inside `aes()`:

```{r, purl=FALSE}
ggplot(data = booksHighUsage,
       aes(x = call_class,
           y = tot_chkout,
           color = subCollection)) +
  geom_point() +
  scale_y_log10()
```

`ggplot()` automatically assigns a unique level of the aesthetic to each unique
value of the variable (this is called *scaling*). Now we reveal indeed that
youth materials make up a large number of high usage items in both the E and the
P class.

Use `fill()` with `geom_bar()` to create a stacked bar plot to visualize
frequency. Again, this reinforced the fact that most of the `E` and `P`
classification are youth materials.

```{r, purl=FALSE}
ggplot(data = booksHighUsage, aes(x = call_class)) +
  geom_bar(aes(fill = subCollection))
```

Stacked bar charts are generally more difficult to read
than side-by-side bars. We can separate the portions of the stacked bar that
correspond to each village and put them side-by-side by using the `position`
argument for `geom_bar()` and setting it to "dodge".

```{r barplot-dodge}
ggplot(data = booksHighUsage, aes(x = call_class)) +
  geom_bar(aes(fill = subCollection), position = "dodge")
```

The order of the classification scale is sorted for "library order."  The audience of library professionals typically prefer an alphabetical arrangement.  However, the x-axis variable is actually categorical.  Categorical data are easier to read when the bars are sorted by frequency.  An easy way to sort by frequency is to use the `fct_infreq()` function from the `forcats` library.   

```{r}
ggplot(data = booksHighUsage, aes(x = fct_infreq(call_class))) +
  geom_bar()
```

Another visualization issue is labeling.  In many cultures, long labels are easier to read horizontally.  Our goal is to flip the x-axis and reorient the x-axis labels into a horizontal presentation.  To accomplish this, flip the axis coordinates with the `coord_flip()` function.  When we flip the axes it's important to reverse the sorted categorical order.  Do this with `forcats::fct_rev()`.

```{r}
ggplot(data = booksHighUsage, aes(x = fct_rev(fct_infreq(call_class)))) +
  geom_bar() +
  coord_flip()
```

### Plotting time series data

Let's calculate number of counts per year for each format for items published
after 1990 and before 2002 in the `booksHighUsage` data frame created above.

First, we use the `ymd()` function from the `lubridate` package to convert our
publication year into a *POSIXct* object. Pass the `truncated = 2` argument as a
way to indicate that the `pubyear` column does not contain month or day. This
means `1990` becomes `1990-01-01`. This will allow us to plot the number of
books per year.

We will do this by calling `mutate()` to create a new variable `pubyear_ymd`.

```{r, purl=FALSE}
booksPlot <- booksPlot %>%
  mutate(pubyear_ymd = ymd(pubyear, truncated = 2))  # convert pubyear to a Date object with ymd()

class(booksPlot$pubyear)  # integer

class(booksPlot$pubyear_ymd)  # Date
```

Next we can use `filter` to remove the `NA` values and get books published
between 1990 and 2003. Notice that we use the `&` as an AND operator to indicate
that the date must fall between that range. We then need to group the data and
count records within each group.

```{r, purl=FALSE}
yearly_counts <- booksPlot %>%
  filter(!is.na(pubyear_ymd),
         pubyear_ymd > "1989-01-01" & pubyear_ymd < "2002-01-01") %>%
  count(pubyear_ymd, subCollection)
```

Time series data can be visualized as a line plot with years on the x axis and counts
on the y axis:

```{r first-time-series, purl=FALSE}
ggplot(data = yearly_counts, mapping = aes(x = pubyear_ymd, y = n)) +
     geom_line()
```

Unfortunately, this does not work because we plotted data for all the sub-collections
together. We need to tell ggplot to draw a line for each sub-collection by modifying
the aesthetic function to include `group = subCollection`:

```{r, purl=FALSE}
ggplot(data = yearly_counts, mapping = aes(x = pubyear_ymd, y = n, group = subCollection)) +
    geom_line()
```

We will be able to distinguish sub-collections in the plot if we add colors
(using `color` also automatically groups the data):

```{r time-series-with-colors, purl=FALSE}
ggplot(data = yearly_counts, mapping = aes(x = pubyear_ymd, y = n, color = subCollection)) +
  geom_line()
```


### Add a third variable with facets

Rather than creating a single plot with side-by-side bars for each
sub-collection, we may want to create multiple plots, where each plot shows the
data for a single sub-collection. This would be especially useful if we had a
large number of sub-collections that we had sampled, as a large number of
side-by-side bars will become more difficult to read.

**`ggplot2`** has a special technique called *faceting* that allows the user to
split one plot into multiple plots based on a factor included in the dataset.

There are two types of `facet` functions:

* `facet_wrap()` arranges a one-dimensional sequence of panels to allow them to cleanly
fit on one page.
* `facet_grid()` allows you to form a matrix of rows and columns of panels.

Both geometries allow to to specify faceting variables specified within `vars()`.
For example, `facet_wrap(facets = vars(facet_variable))` or
`facet_grid(rows = vars(row_variable), cols = vars(col_variable))`.

Here we use `facet_wrap()` to make a time series plot for each subCollection

```{r first-facet, fig.alt = "Three line plots, one each for general collection, juvenile, and K-12 sub-collection materials, showing the relationship of count of books to publication year", purl=FALSE}
ggplot(data = yearly_counts, mapping = aes(x = pubyear_ymd, y = n)) +
    geom_line() +
    facet_wrap(facets = vars(subCollection))
```

We can use `facet_wrap()` as a way of seeing the categories within a variables. Look at the number of formats per
sub-collection. 

```{r purl=FALSE}
ggplot(data = books2, aes(x = fct_rev(fct_infreq(subCollection)))) +
  geom_bar() +
  facet_wrap(~ format, nrow = 2) +
  scale_y_log10() +
  coord_flip() +
  labs(x = "Library subcollection", y = "")
```

While this may not be the most beautiful plot, these kinds of exercises can be
helpful for data exploration. We learn that there are books in all
sub-collections; there are CD-ROMS, serials, images, and microforms in
government documents, and so on. Exploratory plots visually
surface information about your data that are otherwise difficult to parse.
Sometimes they may not meet all the rules for creating beautiful data, but when
you are simply getting to know your data, that's OK.

> ### Challenge
>
> Use the `books2` data to create a bar plot that depicts the number of items in
each sub-collection, faceted by format.
> Add the scale_y_log10() argument to
create a logarithmic scale for easier visibility.
> Add the following theme
argument to tilt the axis text diagonal: `theme(axis.text.x = element_text(angle
= 60, hjust = 1))`
>
> > ## Solution
> > ```{r average-weight-time-series, answer=TRUE, purl=FALSE}
> > ggplot(data = books2, aes(x = subCollection)) +
> > geom_bar() +
> > facet_wrap(vars(format)) +
> > scale_y_log10() +
> > theme(axis.text.x = element_text(angle = 60, hjust = 1))
> > ```
> {: .solution}
{: .challenge}

## **`ggplot2`** themes

Usually plots with white background look more readable when printed.
Every single component of a `ggplot` graph can be customized using the generic
`theme()` function. However, there are pre-loaded themes
available that change the overall appearance of the graph without much effort.

For example, we can change our graph to have a simpler white background
using the `theme_bw()` function:

```{r, purl=FALSE}
#
ggplot(data = yearly_counts, mapping = aes(x = pubyear_ymd, y = n)) +
  geom_line() +
  facet_wrap(facets = vars(subCollection)) +
  theme_bw()
```

In addition to `theme_bw()`, which changes the plot background to white, **`ggplot2`**
comes with several other themes which can be useful to quickly change the look
of your visualization. The complete list of themes is available
at <https://ggplot2.tidyverse.org/reference/ggtheme.html>. `theme_minimal()` and
`theme_light()` are popular, and `theme_void()` can be useful as a starting
point to create a new hand-crafted theme.

The
[ggthemes](https://jrnold.github.io/ggthemes/reference/index.html) package
provides a wide variety of options.
The [**`ggplot2`** extensions website](https://exts.ggplot2.tidyverse.org/) provides a list
of packages that extend the capabilities of **`ggplot2`**, including additional
themes.

## Customization

Take a look at the [**`ggplot2`** cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf), and
think of ways you could improve your plots.

For example, by default, the axes labels on a plot are determined by the name of
the variable being plotted. We can change names of axes to something more
informative than 'pubyear_ymd' and 'n' and add a title to the figure:

```{r, purl=FALSE}
# add labels
ggplot(data = yearly_counts, mapping = aes(x = pubyear_ymd, y = n)) +
  geom_line() +
  facet_wrap(facets = vars(subCollection)) +
  theme_bw() +
    labs(title = "Number of High Usage Books per Year of Publication, by Sub-Collection",
        x = "Year of publication",
        y = "Number of books")
```

Note that it is also possible to change the fonts of your plots. If you are on
Windows, you may have to install
the [**`extrafont`** package](https://github.com/wch/extrafont), and follow the
instructions included in the README for this package.

You can also assign a `theme` to an object in your environment, and pass that
theme to your plot. This can be helpful to keep your `ggplot()` calls less
cluttered. Here we create a `gray_theme` :

```{r}
# create the gray theme
gray_theme <- theme(axis.text.x = element_text(color = "gray20", size = 12, angle = 45, hjust = 0.5, vjust = 0.5),
                    axis.text.y = element_text(color = "gray20", size = 12),
                    text = element_text(size = 16),
                    plot.title = element_text(hjust = 0.5))

# pass the gray theme to a plot
ggplot(data = yearly_counts, mapping = aes(x = pubyear_ymd, y = n)) +
  geom_line() +
  facet_wrap(facets = vars(subCollection)) +
  gray_theme +
  labs(title = "Number of High Usage Books per Year of Publication, \n by Sub-Collection",
        x = "Year of publication",
        y = "Number of books")
```


> ### Challenge
>
> Use the `booksPlot` data to create a plot that depicts how the total number of checkouts
> changes based on year of publication.
>
> First, create a data frame `yearly_checkouts` That meets the following conditions:
> * `filter()` to exclude `NA` values
> * `filter()` between "1989-01-01" and "2002-01-01"
> * `group_by()` the publication year (make sure to use the special Date pubyear value we created)
> * `summarize()` to create a new value `checkouts_sum` that represents the `sum()` of total checkouts per publication year
>
> Then, create a `ggplot` that visualizes the sum of item checkouts by year of
publication. Add one of the themes listed above.
> > ## Solution
> > ```{r}
> > yearly_checkouts <- booksPlot %>%
> >  filter(!is.na(pubyear_ymd),
> >         pubyear_ymd > "1989-01-01" & pubyear_ymd < "2002-01-01") %>%
> >  group_by(pubyear_ymd) %>%
> >  summarize(checkouts_sum = sum(tot_chkout))
> >
> > ggplot(data = yearly_checkouts, mapping = aes(x = pubyear_ymd, y = checkouts_sum)) +
> >  geom_line() +
> >  theme_bw()
> > ```
> {: .solution}
{: .challenge}


## Save and export

After creating your plot, you can save it to a file in your favorite format. The
Export tab in the **Plot** pane in RStudio will save your plots at low
resolution, which will not be accepted by many journals and will not scale well
for posters.

Instead, use the `ggsave()` function, which allows you easily change the
dimension and resolution of your plot by adjusting the appropriate arguments
(`width`, `height` and `dpi`). We have been printing our plot output directly to
the console. To use `ggsave()`, first assign the plot to a variable in your R
environment, such as `yearly_counts_plot`.

Make sure you have the `fig_output/` folder in your working directory.


```{r, eval=FALSE}
yearly_counts_plot <- ggplot(data = yearly_counts, mapping = aes(x = pubyear_ymd, y = n)) +
  geom_line() +
  facet_wrap(facets = vars(subCollection)) +
  gray_theme +
  labs(title = "Number of High Usage Books per Year of Publication, \n by Sub-Collection",
        x = "Year of publication",
        y = "Number of books")

ggsave("fig_output/yearly_counts_plot.png", yearly_counts_plot, width = 15, height = 10)
```


> ## Exercise
>
> With all of this information in hand, please take another five minutes to
> either improve one of the plots generated in this exercise or create a
> beautiful graph of your own. Use the RStudio [**`ggplot2`** cheat sheet](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf)
> for inspiration. Here are some ideas:
>
> * See if you can make the bars white with black outline.
> * Try using a different color palette (see
>   [http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/)).
{: .challenge}


{% include links.md %}
