---
title: "Introduction to R"
keypoints:
- Use the assignment operator <- to assign values to objects. You can now manipulate
  that object in R
- R contains a number of functions you use to do something with your data. Functions
  automate more complicated sets of commands. Many functions are predefined, or can
  be made available by importing R packages
- A vector is a sequence of elements of the same type. All data in a vector must be
  of the same type--character, numeric (or double), integer, and logical. Create vectors
  with c(). Use [ ] to subset values from vectors.
objectives:
- Assign values to objects in R.
- Learn how to name objects.
- Use comments to inform script.
- Solve simple arithmetic operations in R.
- Call functions and use arguments to change their default options.
- Inspect the content of vectors and manipulate their content.
- Subset and extract values from vectors.
- Analyze vectors with missing data.
- 'Define the following terms as they relate to R: object, vector, assign, call, function.'
questions:
- What is an object?
- What is a function and how can we pass arguments to functions?
- How can values be initially assigned to variables of different data types?
- How can a vector be created What are the available data types?
- How does R treat missing values?
- How can we deal with missing values in R?
source: Rmd
teaching: 50
exercises: 30
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("01-")
```


## Creating objects in R

You can get output from R simply by typing math in the console:

```{r, purl=FALSE}
3 + 5
7 * 2   # multiply 7 by 2
sqrt(36) # take the square root of 36
```

However, to do useful and interesting things, we need to assign _values_ to
_objects_. To create an object, we need to give it a name followed by the
assignment operator `<-`, and the value we want to give it:

```{r, purl=FALSE}
time_minutes <- 5        # assign the number 5 to the object time_minutes
```

`<-` is the assignment operator. It assigns values on the right to objects on
the left. Here we are creating a symbol called `time_minutes` and *assigning* it
the numeric value 5. Some R users would say "`time_minutes` *gets* 5."
`time_minutes` is now a *numeric vector* with one element. Or you could say
`time_minutes` is a numeric vector, and the first element is the number 5. 

When you assign something to a symbol, nothing happens in the console, but in
the Environment pane in the upper right, you will notice a new object,
`time_minutes`.

In RStudio, typing <kbd>Alt</kbd> + <kbd>-</kbd> (push <kbd>Alt</kbd> at the
same time as the <kbd>-</kbd> key) will write `<- ` in a single keystroke in a
PC, while typing <kbd>Option</kbd> + <kbd>-</kbd> (push <kbd>Option</kbd> at the
same time as the <kbd>-</kbd> key) does the same in a Mac.

Objects can be given any name such as `x`, `checkouts`, or `isbn`. You want your
object names to be explicit and not too long. Here are some tips for assigning
values:

* **Do not use names of functions that already exist in R:** There are some
names that cannot be used because they are the names of fundamental functions in
R (e.g., `if`, `else`, `for`, see
[here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html)
for a complete list. In general, even if it's allowed, it's best to not use
other function names (e.g., `c`, `T`, `mean`, `data`, `df`, `weights`). If in
doubt, check the help to see if the name is already in use.
* **R is case sensitive**: `age` is different from `Age` and `y` is different
from `Y`.
* **No blank spaces or symbols other than underscores**: R users get around this
in a couple of ways, either through capitalization (e.g. `myData`) or
underscores (e.g. `my_data`). It's also best to avoid dots (`.`) within an
object name as in `my.dataset`. There are many functions in R with dots in their
names for historical reasons, but dots have a special meaning in R (for methods)
and other programming languages.
* **Do not begin with numbers or symbols**: `2x` is not valid, but `x2` is.
* **Be descriptive, but make your variable names short**: It's good practice to
be descriptive with your variable names. If you're loading in a lot of data,
choosing `myData` or `x` as a name may not be as helpful as, say, `ebookUsage`.
Finally, keep your variable names short, since you will likely be typing them in
frequently.


> ## Objects vs. variables
>
> What are known as `objects` in `R` are known as `variables` in many other
> programming languages. Depending on the context, `object` and `variable` can
> have drastically different meanings. However, in this lesson, the two words
> are used synonymously. For more information see:
> [https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Objects](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Objects)
{: .callout}

## Evaluating Expressions

If you now type `time_minutes` into the console, and press Enter on your keyboard, R will
*evaluate* the expression. In this case, R will *print* the elements that are
assigned to `time_minutes` (the number 5). We can do this easily since y only has one
element, but if you do this with a large dataset loaded into R, it will overload
your console because it will print the entire thing. The `[1]` indicates that
the number 5 is the first element of this vector.

When assigning a value to an object, R does not print anything to the console. You
can force R to print the value by using parentheses or by typing
the object name:

```{r, purl=FALSE}
time_minutes <- 5    # doesn't print anything
(time_minutes <- 5)  # putting parenthesis around the call prints the value of y
time_minutes         # so does typing the name of the object
print(time_minutes)  # so does using the print() function.
```

Now that R has `time_minutes` in memory, we can do arithmetic with it. For
instance, we may want to convert it into seconds (60 seconds in 1 minute):

```{r, purl=FALSE}
60 * time_minutes
```

We can also change an object's value by assigning it a new one:

```{r, purl=FALSE}
time_minutes <- 10
60 * time_minutes
```

This overwrites the previous value without prompting you, so be careful! Also,
assigning a value to one object does not change the values of other objects  For
example, let's store the time in seconds in a new object, `time_seconds`:

```{r, purl=FALSE}
time_seconds <- 60 * time_minutes
```

Then change `time_minutes` to 30:

```{r, purl=FALSE}
time_minutes <- 30
```

> ## Exercise
> 
> What do you think is the current content of the object `time_seconds`? 600 or
> 1800?
>
> > ## Solution
> >
> > The value of `time_seconds` is still 600 because you have not
> > re-run the line `time_seconds <- 60 * time_minutes` since
> > changing the value of `time_minutes`.
> {: .solution}
{: .challenge}

> ## Exercise
>
> Create two variables `my_length` and `my_width` and assign them any numeric values you want.
> Create a third variable `my_area` and give it a value based on the the multiplication of `my_length` and `my_width`.
> Show that changing the values of either `my_length` and `my_width` does not affect the value of `my_area`.
> 
> > ## Solution
> > 
> > ```{r}
> > my_length <- 2.5
> > my_width <- 3.2
> > my_area <- my_length * my_width
> > area
> > # change the values of my_length and my_width
> > my_length <- 7.0
> > my_width <- 6.5
> > # the value of my_area isn't changed
> > my_area
> > ```
> > 
> {: .solution}
{: .challenge}

## Comments

All programming languages allow the programmer to include comments in their
code. To do this in R we use the `#` character. Anything to the right of the `#`
sign and up to the end of the line is treated as a comment and will not be
evaluated by R. You can start lines with comments or include them after any code
on the line.

Comments are essential to helping you remember what your code does, and
explaining it to others. Commenting code, along with documenting how data is
collected and explaining what each variable represents, is essential to
reproducible research. See the Software Carpentry lesson on [R for Reproducible
Scientific Analysis](https://swcarpentry.github.io/r-novice-gapminder/).

```{r}
time_minutes <- 5    # time in minutes
time_seconds <- 60 * time_minutes	# convert to seconds
time_seconds				# print time in seconds
```

RStudio makes it easy to comment or uncomment a paragraph: after selecting the
lines you  want to comment, press at the same time on your keyboard
<kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>C</kbd>. If you only want to comment
out one line, you can put the cursor at any location of that line (i.e. no need
to select the whole line), then press <kbd>Ctrl</kbd> + <kbd>Shift</kbd> +
<kbd>C</kbd>.

## Removing objects from the environment
To remove an object from your R environment, use the `rm()` function. Remove
multiple objects with `rm(list = c("add", "objects", "here))`, adding the
objects in `c()` using quotation marks. To remove all objects, use `rm(list =
ls())` or click the broom icon in the Environment Pane, next to "Import
Dataset."

```{r, purl=FALSE, eval=FALSE}
x <- 5
y <- 10
z <- 15
rm(x)  # remove x
rm(list =c("y", "z"))  # remove y and z
rm(list = ls())  # remove all objects
```


## Functions and their arguments

R is a "functional programming language," meaning it contains a number of
*functions* you use to do something with your data. Functions are "canned
scripts" that automate more complicated sets of commands. Many functions are
predefined, or can be made available by importing R *packages* as we saw in the
"Before We Start" lesson.

*Call* a function on a variable by entering the function into the console,
followed by parentheses and the variables. A function usually gets one or more
inputs called *arguments*. For example, if you want to take the sum of 3 and 4,
you can type in `sum(3, 4)`. In this case, the arguments must be a number, and
the return value (the output) is the sum of those numbers. An example of a
function call is:


```{r, eval=FALSE, purl=FALSE}
sum(3, 4)
```

The function `is.function()` will check if an argument is a function in R. If it
is a function, it will print `TRUE` to the console.

Functions can be nested within each other. For example, `sqrt()` takes the
square root of the number provided in the function call. Therefore you can run
`sum(sqrt(9), 4)` to take the sum of the square root of 9 and add it to 4. 

Typing a question mark before a function will pull the help page up in the
Navigation Pane in the lower right. Type `?sum` to view the help page for the
`sum` function. You can also call `help(sum)`. This will provide the description
of the function, how it is to be used, and the arguments.

In the case of `sum()`, the ellipses `. . .` represent an unlimited number of
numeric elements. 

```{r, eval=FALSE, purl=FALSE}
is.function(sum)        # check to see if sum() is a function
sum(3, 4, 5, 6, 7)      # sum takes an unlimited number (. . .) of numeric elements
```

### Arguments

Some functions take arguments which may either be specified by the user, or, if
left out, take on a *default* value. However, if you want something specific,
you can specify a value of your choice which will be used instead of the
default. This is called *passing* an argument to the function.

For example, `sum()` takes the argument option `na.rm`. If you check the help
page for sum (call `?sum`), you can see that `na.rm` requires a logical
(`TRUE/FALSE`) value specifying whether `NA` values (missing data) should be
removed when the argument is evaluated.

By default, `na.rm` is set to `FALSE`, so evaluating a sum with missing
values will return `NA`:

```{r, eval=TRUE, purl=FALSE}
sum(3, 4, NA)                # 
```

Even though we do not see the argument here, it is operating in the background,
as the `NA` value remains. 3 + 4 + `NA` is `NA`.

But setting the argument `na.rm` to `TRUE` will remove the `NA`:

```{r, eval=TRUE, purl=FALSE}
sum(3, 4, NA, na.rm = TRUE)
```

It is very important to understand the different arguments that functions take,
the values that can be added to those functions, and the default arguments.
Arguments can be anything, not only `TRUE` or `FALSE`, but also other objects.
Exactly what each argument means differs per function, and must be looked up in
the documentation.

It's good practice to put the non-optional arguments first in your function
call, and to specify the names of all optional arguments.  If you don't, someone
reading your code might have to look up the definition of a function with
unfamiliar arguments to understand what you're doing.

> ## Exercise
>
> Type in `?round` at the console and then look at the output in the Help pane.
> What other functions exist that are similar to `round`?
> How do you use the `digits` parameter in the round function?
{: .challenge}



## Vectors and data types

```{r, echo=FALSE, purl=TRUE}
### Vectors and data types
```

A vector is the most common and basic data type in R, and is pretty much the
workhorse of R. A vector is a sequence of elements of the same *type*. Vectors
can only contain "homogenous" data--in other words, all data must be of the same
type. The type of a vector determines what kind of analysis you can do on it.
For example, you can perform mathematical operations on `numeric` objects, but
not on `character` objects.

We can assign a series of values to a vector using the `c()` function. `c()`
stands for combine. If you read the help files for `c()` by calling `help(c)`,
you can see that it takes an unlimited `. . .` number of arguments.

For example we can create a vector of checkouts for a collection of books and
assign it to a new object `checkouts`:

```{r, purl=FALSE}
checkouts <- c(25, 15, 18)
checkouts
```

A vector can also contain characters. For example, we can have
a vector of the book titles (`title`) and authors (`author`):

```{r, purl=FALSE}
title <- c("Macbeth","Dracula","1984")
```

The quotes around "Macbeth", etc. are essential here. Without the quotes R will
assume there are objects called `Macbeth` and `Dracula` in the environment. As
these objects don't yet exist in R's memory, there will be an error message.

There are many functions that allow you to inspect the content of a vector.
`length()` tells you how many elements are in a particular vector:

```{r, purl=FALSE}
length(checkouts)  # print the number of values in the checkouts vector
```

An important feature of a vector, is that all of the elements are the same type
of data. The function `class()` indicates the class (the type of element) of an
object:

```{r, purl=FALSE}
class(checkouts)
class(title)
```

Type `?str` into the console to read the description of the `str` function. You
can call `str()` on an R object to compactly display information about it,
including the data type, the number of elements, and a printout of the first few
elements.

```{r, purl=FALSE}
str(checkouts)
str(title)
```

You can use the `c()` function to add other elements to your vector:

```{r, purl=FALSE}
author <- "Stoker"
author <- c(author, "Orwell") # add to the end of the vector
author <- c("Shakespeare", author)
author
```

In the first line, we create a character vector `author` with a single value
`"Stoker"`. In the second line, we add the value `"Orwell"` to it, and save the
result back into `author`. Then we add the value `"Shakespeare"` to the
beginning, again saving the result back into `author`.

We can do this over and over again to grow a vector, or assemble a dataset.
As we program, this may be useful to add results that we are collecting or
calculating.

An **atomic vector** is the simplest R **data type** and is a linear vector of a
single type. Above, we saw 2 of the 6 main **atomic vector** types  that R uses:
`"character"` and `"numeric"` (or `"double"`). These are the basic building
blocks that all R objects are built from. The other 4 **atomic vector** types
are:

* `"logical"` for `TRUE` and `FALSE` (the boolean data type)
* `"integer"` for integer numbers (e.g., `2L`, the `L` indicates to R that it's an integer)
* `"complex"` to represent complex numbers with real and imaginary parts (e.g.,
`1 + 4i`) and that's all we're going to say about them
* `"raw"` for bitstreams that we won't discuss further

You can check the type of your vector using the `typeof()` function and
inputting your vector as the argument.

Vectors are one of the many **data structures** that R uses. Other important
ones are lists (`list`), matrices (`matrix`), data frames (`data.frame`),
factors (`factor`) and arrays (`array`).

> ## Exercise
>
>
> We’ve seen that atomic vectors can be of type character, numeric (or double),
>   integer, and logical. But what happens if we try to mix these types in a
>   single vector?
>
> > ## Solution
> >
> > R implicitly converts them to all be the same type.
> {: .solution}
>
> What will happen in each of these examples? (hint: use `typeof()`
> to check the data type of your objects):
>
>  ```{r, results="hide"}
>  num_char <- c(1, 2, 3, "a")
>  num_logical <- c(1, 2, 3, TRUE)
>  char_logical <- c("a", "b", "c", TRUE)
>  tricky <- c(1, 2, 3, "4")
>  ```
>
>
> Why do you think it happens?
>
> > ## Solution
> >
> >  Vectors can be of only one data type. R tries to
> > convert (coerce) the content of this vector to find a "common
> > denominator" that doesn't lose any information.
> {: .solution}
>
>
> How many values in `combined_logical` are `"TRUE"` (as a character) in the
> following example:
>
> ```{r, results="hide"}
> num_logical <- c(1, 2, 3, TRUE)
> char_logical <- c("a", "b", "c", TRUE)
> combined_logical <- c(num_logical, char_logical)
>  ```
> &nbsp;
>
> > ## Solution
> >
> > Only one. There is no memory of past data types, and the coercion
> > happens the
> > first time the vector is evaluated. Therefore, the `TRUE` in
> > `num_logical`
> > gets converted into a `1` before it gets converted into `"1"` in
> > `combined_logical`.
> {: .solution}
>
>
{: .challenge}

You've probably noticed that objects of different types get converted into a
single, shared type within a vector. In R, we call converting objects from one
class into another class _coercion_. These conversions happen according to a
hierarchy, whereby some types get preferentially coerced into other types. This
hierarchy is: logical < integer < numeric < complex < character < list.

You can also *coerce* a vector to be a specific data type with `as.character()`,
`as.logical()`, `as.numeric()`, etc. For example, to coerce a number to a
character:

```{r, purl=FALSE}
x <- as.character(200)
```

We can test this in a few ways: if we print `x` to the console, we see quotation
marks around it, letting us know it is a character:

```{r, purl=FALSE}
x
```

We can also call `class()`

```{r, purl=FALSE}
class(x)
```

And if we try to add a number to `x`, we will get an error message `non-numeric
argument to binary operator`--in other words, x is `non-numeric` and cannot be
added to a number.

```{r, purl=FALSE,eval=FALSE}
x + 5
```

## Subsetting vectors

If we want to subset (or extract) one or several values from a vector, we must
provide one or several indices in square brackets. For this example, we will use
the `state` data, which is built into R and includes data related to the 50
states of the U.S.A. Type `?state` to see the included datasets. `state.name` is
a built in vector in R of all U.S. states:

```{r, results='show', purl=FALSE}
state.name
state.name[1]
```

You can use the `:` colon to create a vector of consecutive numbers.
```{r, results='show', purl=FALSE}
state.name[1:5] 
```

If the numbers are not consecutive, you must use the `c()` function:
```{r, results='show', purl=FALSE}
state.name[c(1, 10, 20)]
```

We can also repeat the indices to create an object with more elements than the
original one:

```{r, results='show', purl=FALSE}
state.name[c(1, 2, 3, 2, 1, 3)]
```

R indices start at 1. Programming languages like Fortran, MATLAB, Julia, and R
start counting at 1, because that's what human beings typically do. Languages in
the C family (including C++, Java, Perl, and Python) count from 0 because that's
simpler for computers to do.

### Conditional subsetting

Another common way of subsetting is by using a logical vector. `TRUE` will
select the element with the same index, while `FALSE` will not:

```{r, results='show', purl=FALSE}
five_states <- state.name[1:5]
five_states[c(TRUE, FALSE, TRUE, FALSE, TRUE)]
```

Typically, these logical vectors are not typed by hand, but are the output of
other functions or logical tests. `state.area` is a vector of state areas in
square miles. We can use the `<` operator to return a logical vector with TRUE
for the indices that meet the condition:

```{r, results='show', purl=FALSE}
state.area < 10000
state.area[state.area < 10000]
```

The first expression gives us a logical vector of length 50, where `TRUE`
represents those states with areas less than 10,000 square miles. The second
expression subsets `state.name` to include only those names where the value is
`TRUE`.

You can also specify character values. `state.region` gives the region that each
state belongs to:
```{r, results='show', purl=FALSE}
state.region == "Northeast"
state.name[state.region == "Northeast"]
```

Again, a `TRUE/FALSE` index of all 50 states where the region is the Northeast,
followed by a subset of `state.name` to return only those `TRUE` values.

Sometimes you need to do multiple logical tests (think Boolean logic). You can
combine multiple tests using `|` (at least one of the conditions is true, OR) or
`&` (both conditions are true, AND). Use `help(Logic)` to read the help file.

```{r logical, comment=NA, results='asis', echo=F, eval=FALSE}
multiple <- data.frame("operator" = c("&", "|", "!", "any", "all")
                       , "function" = c("boolean AND", "boolean OR", "boolean NOT", "ANY true", "ALL true")
                       , stringsAsFactors = F)

kable(multiple) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```


```{r, results='show', purl=FALSE}
state.name[state.area < 10000 | state.region == "Northeast"]
state.name[state.area < 10000 & state.region == "Northeast"]
```

The first result includes both states with fewer than 10,000 sq. mi. and all
states in the Northeast. New York, Pennsylvania, Delaware and Maine have areas
with greater than 10,000 square miles, but are in the Northeastern U.S. Hawaii
is not in the Northeast, but it has fewer than 10,000 square miles. The second
result includes only states that are in the Northeast and have fewer than 10,000
sq. mi.

R contains a number of operators you can use to compare values. Use
`help(Comparison)` to read the R help file. Note that **two equal signs** (`==`)
are used for evaluating equality (because one equals sign (`=`) is used for
assigning variables).

```{r logical1, comment=NA, results='asis', echo=F, eval=FALSE}
operators <- tibble("operator" = c("<", ">", "==", "<=", ">=", "!=", "%in%", "is.na", "!is.na"),
                    "function" = c("Less Than", "Greater Than", "Equal To", "Less Than or Equal To", "Greater Than or Equal To", "Not Equal To", "Has A Match In", "Is NA", "Is Not NA"))

kable(operators) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

A common task is to search for certain strings in a vector.  One could use the
"or" operator `|` to test for equality to multiple values, but this can quickly
become tedious. The function `%in%` allows you to test if any of the elements of
a search vector are found:

```{r, results='show', purl=FALSE}
west_coast <- c("California", "Oregon", "Washington")
state.name[state.name == "California" | state.name == "Oregon" | state.name == "Washington"]
state.name %in% west_coast
state.name[state.name %in% west_coast]
```

## Missing data

As R was designed to analyze datasets, it includes the concept of missing data
(which is uncommon in other programming languages). Missing data are represented
in vectors as `NA`. R functions have special actions when they encounter `NA.`

When doing operations on numbers, most functions will return `NA` if the data
you are working with include missing values. This feature makes it harder to
overlook the cases where you are dealing with missing data. As we saw above, you
can add the argument `na.rm=TRUE` to calculate the result while ignoring the
missing values.

```{r, purl=FALSE}
rooms <- c(2, 1, 1, NA, 4)
mean(rooms)
max(rooms)
mean(rooms, na.rm = TRUE)
max(rooms, na.rm = TRUE)
```

If your data include missing values, you may want to become familiar with the
functions `is.na()`, `na.omit()`, and `complete.cases()`. See below for
examples.


```{r, purl=FALSE}
## Use any() to check if any values are missing
any(is.na(rooms))

## Use table() to tell you how many are missing vs. not missing
table(is.na(rooms))

## Identify those elements that are not missing values.
complete.cases(rooms)

## Identify those elements that are missing values.
is.na(rooms)

## Extract those elements that are not missing values.
rooms[complete.cases(rooms)]
```

You can also use `!is.na(rooms)`, which is exactly the same as
`complete.cases(rooms)`. The exclamation mark indicates logical negation.

```{r, purl=FALSE}
!c(TRUE, FALSE)
```


How you deal with missing data in your analysis is a decision you will have to
make--do you remove it entirely? Do you replace it with zeros? That will depend
on your own methodological questions.

> ## Exercise
>
> 1. Using this vector of rooms, create a new vector with the NAs removed.
>
>     ```r
>     rooms <- c(1, 2, 1, 1, NA, 3, 1, 3, 2, 1, 1, 8, 3, 1, NA, 1)
>     ```
> 2. Use the function `median()` to calculate the median of the `rooms` vector.
>
> 3. Use R to figure out how many households in the room variable have more than 2 rooms.
>
> > ## Solution
> > ```{r, answer=TRUE}
> > rooms <- c(1, 2, 1, 1, NA, 3, 1, 3, 2, 1, 1, 8, 3, 1, NA, 1)
> > rooms_no_na <- rooms[!is.na(rooms)]
> > # or
> > rooms_no_na <- na.omit(rooms)
> > # 2.
> > median(rooms, na.rm = TRUE)
> > # 3.
> > rooms_above_2 <- rooms_no_na[rooms_no_na > 2]
> > length(rooms_above_2)
> > ```
> {: .solution}
{: .challenge}

Now that we have learned how to write scripts, and the basics of R's data
structures, we are ready to start working with the library catalog dataset and learn about data frames.


{% include links.md %}
