---
title: Reproducible Reports with Quarto
teaching: 45
exercises: 15
source: Rmd
---

::::::::::::::::::::::::::::::::::::::: objectives

- Describe the value of reproducible reporting.
- Create a new Quarto document (`.qmd`) in RStudio.
- Use Markdown syntax to format text.
- Create and run code chunks within a Quarto document.
- Render a Quarto document to an HTML report.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How can I combine my code, results, and narrative into a single document?
- How can I automatically update my reports when my data changes?
- What is Quarto and how does it differ from a standard R script?

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, include=FALSE}
source("files/download_data.R")
library(tidyverse)
# Read raw data and apply cleaning steps from previous episodes
books2 <- read_csv("data/books.csv") %>%
  rename(
    title = X245.ab,
    author = X245.c,
    callnumber = CALL...BIBLIO.,
    isbn = ISN,
    pubyear = X008.Date.One,
    subCollection = BCODE1,
    format = BCODE2,
    location = LOCATION,
    tot_chkout = TOT.CHKOUT,
    loutdate = LOUTDATE,
    subject = SUBJECT,
    callnumber2 = CALL...ITEM.
  ) %>%
  mutate(
    pubyear = as.integer(pubyear),
    subCollection = recode(subCollection,
      "-" = "general collection",
      u = "government documents",
      r = "reference",
      b = "k-12 materials",
      j = "juvenile",
      s = "special collections",
      c = "computer files",
      t = "theses",
      a = "archives",
      z = "reserves"
    ),
    format = recode(format,
      a = "book",
      e = "serial",
      w = "microform",
      s = "e-gov doc",
      o = "map",
      n = "database",
      k = "cd-rom",
      m = "image",
      "5" = "kit/object",
      "4" = "online video"
    )
  )
```

## Introduction to Reproducible Reporting

So far, we have been writing code in `.R` scripts. This is excellent for data analysis, but what happens when you need to share your findings with a colleague or a library director? You might copy a plot into a Word document or an email, then type out your interpretation. 

But what if the data changes next month? You would have to re-run your script, re-save the plot, copy it back into Word, and update your text. This manual process is prone to errors and tedious.

**Quarto** allows you to combine your code, its output (plots, tables), and your narrative text into a single document. When you "render" the document, R runs the code and produces a polished report (HTML, PDF, or Word) automatically.

## Creating a Quarto Document

To create a new Quarto document in RStudio:

1. Click the **File** menu.
2. Select **New File** > **Quarto Document...**
3. In the dialog box, give your document a **Title** (e.g., "Library Usage Report") and enter your name as **Author**.
4. Ensure **HTML** is selected as the output format.
5. Click **Create**.

RStudio will open a new file with some example content. Notice the file extension is `.qmd`.

:::::::::::::::::::::::::::::::::::::::::  callout

## Quarto vs. RMarkdown

If you have used R before, you might be familiar with RMarkdown (`.Rmd`). Quarto (`.qmd`) is the next-generation version of RMarkdown. It works very similarly but supports more languages (like Python and Julia) and has better features for scientific publishing.

::::::::::::::::::::::::::::::::::::::::: 

## Anatomy of a Quarto Document

A Quarto document has three main parts:

### 1. The YAML Header

At the very top, enclosed between two lines of `---`, is the **YAML Header**. This contains metadata about the document.

```yaml
---
title: "Library Usage Report"
author: "Your Name"
format: html
---
```

### 2. Markdown Text

The white space is where you write your narrative. You use **Markdown** syntax to format text. 

- `**Bold**` for **bold text**
- `*Italics*` for *italics*
- `# Heading 1` for a main title
- `## Heading 2` for a section title
- `- List item` for bullet points

### 3. Code Chunks

Code chunks are where your R code lives. They start with ` ```{r} ` and end with ` ``` `.

````
```{r}
# This is a code chunk
summary(cars)
```
````

You can insert a new chunk by clicking the **+C** button in the editor toolbar, or by pressing <kbd>Ctrl</kbd>+<kbd>Alt</kbd>+<kbd>I</kbd> (Windows/Linux) or <kbd>Cmd</kbd>+<kbd>Option</kbd>+<kbd>I</kbd> (Mac).

## Your First Report

Let's clean up the example file and create a report using our `books` data.

1.  Delete everything in the file *below* the YAML header.
2.  Add a new **setup** code chunk to load our libraries and prepare the data.

```{{r}}
#| label: setup
#| include: false

library(tidyverse)

# Load data and rename columns for clarity
books2 <- read_csv("data/books.csv") %>%
  rename(
    subCollection = BCODE1,
    tot_chkout = TOT.CHKOUT,
    format = BCODE2
  ) %>%
  mutate(
    subCollection = recode(subCollection, 
      "-" = "general collection", 
      j = "juvenile", 
      b = "k-12 materials"
    )
  )
```

:::::::::::::::::::::::::::::::::::::::::  callout

## Chunk Options

Notice the lines starting with `#|`. These are **chunk options**. 
- `#| label: setup` gives the chunk a name.
- `#| include: false` runs the code but hides the code and output from the final report. This is great for loading data silently.

::::::::::::::::::::::::::::::::::::::::: 

### Adding Analysis

Now, let's add a section header and some text.

```markdown
## High Usage Items

We are analyzing items with more than 10 checkouts to understand circulation patterns across sub-collections.
```

Next, insert a new code chunk and paste the plotting code we developed in the previous episode (ggplot2).

````
```{r}
#| label: plot-high-usage
#| echo: false

# Filter for high usage
booksHighUsage <- books2 %>%
  filter(!is.na(tot_chkout),
         tot_chkout > 10)

# Create the plot
ggplot(data = booksHighUsage, 
       aes(x = subCollection, y = tot_chkout)) +
  geom_boxplot(alpha = 0) +
  geom_jitter(alpha = 0.5, color = "tomato") +
  scale_y_log10() +
  labs(title = "Distribution of Checkouts by Sub-Collection",
       x = "Sub-Collection",
       y = "Total Checkouts (Log Scale)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
````

Setting `#| echo: false` will display the *plot* in the report, but hide the R *code* that generated it. This is often preferred for reports intended for non-coders.

## Rendering the Document

Now comes the magic. Click the **Render** button (blue arrow icon) at the top of the editor pane.

RStudio will:
1.  Run all your code chunks from scratch.
2.  Generate the plots and results.
3.  Combine them with your text.
4.  Create a new file named `library_usage_report.html` in your project folder.
5.  Open a preview of the report.

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge: Add a Summary Table

1.  Add a new header `## Summary Statistics` to your Quarto document.
2.  Insert a new code chunk.
3.  Write code to calculate the mean checkouts per format (Hint: use `group_by(format)` and `summarize()`).
4.  Render the document again to see your new table included in the report.

:::::::::::::::  solution

## Solution

Add this to your document:

```markdown
## Summary Statistics

The table below shows the average checkouts for each item format.
```

````
```{r}
#| label: summary-table

books2 %>%
  group_by(format) %>%
  summarize(mean_checkouts = mean(tot_chkout, na.rm = TRUE)) %>%
  arrange(desc(mean_checkouts))
```
````

Render the document to see the updated report.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Why This Matters

By using Quarto, your report is now **reproducible**. 

If you download a new version of `books.csv` next month:
1.  Save it to your `data/` folder.
2.  Open your Quarto document.
3.  Click **Render**.

Your report will automatically update with the new data, creating a fresh plot and table without you having to copy-paste a single thing.

:::::::::::::::::::::::::::::::::::::::: keypoints

- **Quarto** allows you to mix code and text to create reproducible reports.
- Use the **YAML header** to configure document metadata like title and output format.
- **Code chunks** run R code and can display or hide input/output using options like `#| echo: false`.
- **Rendering** the document executes the code and produces the final output (HTML, PDF, etc.).
- This workflow saves time and reduces errors when reporting on data that changes over time.

::::::::::::::::::::::::::::::::::::::::::::::::::
